---
title: "Clustering_Review"
author: "Foo Lin Geng"
date: "Term 6, 2020"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 3
    number_sections: true
---

First, we prepare the working environment.
```{r}
rm(list=ls()) # Clear the environment
# setwd("...")  # Setup the working directory
```


# Pollution dataset

This datasets was proposed in McDonald, G.C. and Schwing, R.C. (1973) 'Instabilities of regression estimates relating air pollution to mortality', Technometrics, vol.15, 463-482. It contains 16 attributes describing 60 different pollution scenarios. The attributes are the following: 

1) PRECReal: Average annual precipitation in inches 
2) JANTReal: Average January temperature in degrees F 
3) JULTReal: Same for July 
4) OVR65Real: of 1960 SMSA population aged 65 or older 
5) POPNReal: Average household size 
6) EDUCReal: Median school years completed by those over 22 
7) HOUSReal: of housing units which are sound and with all facilities 
8) DENSReal: Population per sq. mile in urbanized areas, 1960 
9) NONWReal: non-white population in urbanized areas, 1960 
10) WWDRKReal: employed in white collar occupations 
11) POORReal: of families with income less than $3000 
12) HCReal: Relative hydrocarbon pollution potential 
13) NOXReal: Same for nitric oxides 
14) SOReal: Same for sulphur dioxide 
15) HUMIDReal: Annual average % relative humidity at 1pm 
16) MORTReal: Total age-adjusted mortality rate per 100,000

```{r}
pollution <- read.csv("pollution.csv")
#str(pollution)
```

A quick look at the dataframe:

```{r}
head(pollution) # First part of the dataframe
summary(pollution) # Summary of the data
```


# Hierarchical clustering

First, use the function `dist` to calculate the distance between entries (rows) using all columns of data. Then, execute hierarchical clustering using Ward's distance.

```{r}
distances <- dist(pollution, method="euclidean")
clusterpollution1 <- hclust(distances, method="ward.D2")
```

Plot the dendrogram.

```{r}
plot(clusterpollution1) 
```

Let's now cut the dendrogram into 4 clusters and compute the average value across the clusters for the MORTReal variable.

```{r}
clusterGroups1 <- cutree(clusterpollution1, k=4)
tapply(pollution$MORTReal, clusterGroups1, mean) 
```

Create a matrix "Centroids" containing the centroids of each cluster, where rows denote variables and columns indicate clusters

```{r}
Centroid <- matrix(0,nrow=16,ncol=4)
for(i in 1:16)
{Centroid[i,] <- tapply(pollution[,i], clusterGroups1, mean)}
rownames(Centroid) <- colnames(pollution)[1:16]
Centroid
```

Looking at `Centroid`, we can observe that DENSReal and MORTReal are the variables with the most variation, and the data is clustered primarily according to those two variables.

Let's plot at the clustering result along DENSReal and MORTReal, with the color differentiating the clusters:
```{r}
plot(pollution$DENSReal,pollution$MORTReal, col=clusterGroups1, pch=19)
```

This is not natural.
That shows the importance of scaling. We need to scale our data for better performance!
Scale the data using the scale() function and repeat the steps above to display the dendrogram.

```{r}
pollution_scaled = data.frame(scale(pollution)) 
colnames(pollution_scaled) = names(pollution)
distances_scaled <- dist(pollution_scaled, method="euclidean")
clusterpollution2 <- hclust(distances_scaled, method="ward.D2")
plot(clusterpollution2) 
```


Let's  do the same as before: cut the dendrogram into 4 clusters and plot it along DENSReal and MORTReal, with points colored according to their group.
```{r}
clusterGroups2 <- cutree(clusterpollution2, k=4)
plot(pollution_scaled$DENSReal,pollution_scaled$MORTReal, col=clusterGroups2, pch=19)
```

We observe that the clusters show much more overlap than before as expected. This shows the importance of scaling the data!

Now, let's plot 3 different plots (all with the points colored according to their group):
1) HCReal vs MORTReal
2) NOXReal vs MORTReal
3) SOReal vs MORTReal
```{r}
plot(pollution_scaled$HCReal,pollution_scaled$MORTReal, col=clusterGroups2, pch=19)
plot(pollution_scaled$NOXReal,pollution_scaled$MORTReal, col=clusterGroups2, pch=19)
plot(pollution_scaled$SOReal,pollution_scaled$MORTReal, col=clusterGroups2, pch=19)
```

# K-means clustering

First set the seed.

```{r}
set.seed(1)
```

Perform a K-means clustering using k=4 and 20 random initial configurations. Print the total within cluster sum of squares.

```{r}
clusterpollution3 <- kmeans(pollution_scaled,centers=4,nstart=20)
clusterpollution3$tot.withinss 
```

Let's now vary the value of k and plot the error in the cluster as a function of k:

```{r}
set.seed(1)
fit <- 0
for(k in 1:15)
{clusterpollution4 <- kmeans(pollution_scaled, centers=k, nstart=20)
fit[k] <- clusterpollution4$tot.withinss}
plot(1:15,fit)
```

<!-- Let's see the corresponding clusters when there are 4 clusters: -->

<!-- ```{r} -->
<!-- Centroid2 <- matrix(0,nrow=16,ncol=4) -->
<!-- for(i in 1:16) -->
<!-- {Centroid2[i,] <- tapply(pollution_scaled[,i], clusterpollution3$cluster, mean)} -->
<!-- rownames(Centroid2) <- colnames(pollution_scaled)[1:16] -->
<!-- Centroid2 -->
<!-- ``` -->

Now, let's plot 3 different plots (all with points colored according to their group):
1) HCReal vs MORTReal
2) NOXReal vs MORTReal
3) SOReal vs MORTReal
```{r}
plot(pollution_scaled$HCReal,pollution_scaled$MORTReal, col=clusterpollution3$cluster, pch=19)
plot(pollution_scaled$NOXReal,pollution_scaled$MORTReal, col=clusterpollution3$cluster, pch=19)
plot(pollution_scaled$SOReal,pollution_scaled$MORTReal, col=clusterpollution3$cluster, pch=19)
```

The results look very similar to the one obtained from hierarchical clustering.

# BONUS: PCA on top of K-means
## Please do not panic. This is just cool, and good to know.

Let's use PCA in order to better visualize the classes!

```{r}
pr.out<-prcomp(pollution_scaled,scale=F)
#summary(pr.out) 
pve<-pr.out$sdev^2/sum(pr.out$sdev^2)
cpve<-cumsum(pve)
plot(cpve,xlab="Principal components",type="l",ylim=c(0,1))
```


```{r}
library(factoextra)
fviz_pca_biplot(pr.out, label = "var", habillage=clusterpollution3$cluster,addEllipses=TRUE, ellipse.level=0.95)
```

We observe one group that has more education, more white-collar jobs and  better housing units (also higher HC,NOX pollution).
Another group has more poorer families, higher mortaility rates, higher non-white populaion.
It seems like, in this dataset, the potential for pollution (in the area) is not linked to mortality rates.

For convenience:
1) PRECReal: Average annual precipitation in inches 
2) JANTReal: Average January temperature in degrees F 
3) JULTReal: Same for July 
4) OVR65Real: of 1960 SMSA population aged 65 or older 
5) POPNReal: Average household size 
6) EDUCReal: Median school years completed by those over 22 
7) HOUSReal: of housing units which are sound and with all facilities 
8) DENSReal: Population per sq. mile in urbanized areas, 1960 
9) NONWReal: non-white population in urbanized areas, 1960 
10) WWDRKReal: employed in white collar occupations 
11) POORReal: of families with income less than $3000 
12) HCReal: Relative hydrocarbon pollution potential 
13) NOXReal: Same for nitric oxides 
14) SOReal: Same for sulphur dioxide 
15) HUMIDReal: Annual average % relative humidity at 1pm 
16) MORTReal: Total age-adjusted mortality rate per 100,000






