---
title: "Week 1 Lecture 1: R Basics"
author: "Stefano Galelli"
date: "Term 5, 2022"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: yes
    toc_depth: 2
    number_sections: yes
---

# Preliminaries

Check [bookdown](http://bookdown.org) for R Markdown and other information, tutorial on R and R studio.

## R environment in R Markdown

Use `ctrl+alt+i` to create the R environment

## Current directory

```{r}
getwd()
```


## Help function
```{r}
help(getwd)
?getwd
```

## Set directory
Setting the current working directory
```{r}
setwd("/Volumes/DATA/Dropbox/The Analytics Edge -- Summer 2022/Week 1/R code/lec1/ver2022")
```

## List the files in a directory

```{r}
dir("/Volumes/DATA/Dropbox/The Analytics Edge -- Summer 2022/Week 1/R code/lec1/")
```

## List objects 
Listing objects stored in the R workspace

```{r}
ls()
```

## Variable assignments

Assign a number to a variable.

```{r}
x<-45
x
x=45 #Alternative to assigning a number to a variable. In general '<-' is preferred for assignment.
x
```


## Common functions 
Exponential, inverse, power, addition.
```{r}
exp(x)
1/x
x^2
x+5
y<-x+5
y
z<-x+y
z
```

    
## Remove variables from the workspace

```{r}
ls()
rm(ct)
ls()
rm(list=ls())
ls()
```

# Numbers and vectors


## Concatenation
Concatenate (combine) numbers to form a vector.
```{r}
x<-c(1,-2,pi,55,45)
x
```


## Accessing elements
Access specific elements of the vector.
```{r}
x[4]
x[3]
```

## Applying operations to a vector 
Term by term inverse, concatenate vectors, exponentiation.
```{r}
1/x
y<-c(x,0,x)
y
x+y # be carefull while adding vectors
x
y
```

You can overload the sum operator by recycling the shorter vector - mathematically adding vectors of different sizes are not permitted. 


## Maximum and minimum
Finding the maximum and minimum elements and identifying the location (index) of the first max and all max.
```{r}
max(x)
min(y)
y
which.max(x)
which.min(y)
y
which(y==min(y))
```


## Other operations 
  - sum, product, mean, variance, standard deviation
```{r}
sum(x)
cumsum(x)
prod(x)
mean(x)
sd(x)
var(x)
summary(x)
```


For a vector, *summary* it provides a six number summary including min, max, mean, 1st quartile, median and 3rd quartile. This can be used with other objects too.

## Maximum entry in a concatenated vector

```{r}
max(x,y)
```

## Parallel maximum
```{r}
x1<- 3:8
x2<- 8:3
x1
x2
max(x1,x2)
pmax(x1,x2)
pmax(x,y)
```


Parallel maximum returns a vector of length equal to the longest argument that contains in each element, the largest element in that position of any of the vector.

## Generating vectors using a variety of commands

```{r}
x<- -3:8
x

y<-seq(-3,8,0.2)
y

z<-seq(-3,8, length=5)
z

rep(x, times=3)
rep(x, each=3)

```

## Return logical vector based on the check
```{r}
x
x>1
```


## Dealing with missing entries 

```{r}
is.na(x)
y<-c(x,NA)

y
is.na(y)
na.omit(y)
?na.omit
```

## Differences in assignments
   Using <- vs = vs ==.
```{r}
exp(1:5)
#exp(a=1:5)
exp(a<-1:5)
a
exp(a==1:5)
```

   

# Classes or types

R stores objects in a variety of classes - *numeric, integer, character, logical, list, matrix, dataframe* and has logical overriding operations when you convert from one class to another.

## Variety of classes
```{r}
z<-0:9
z
class(z)

z1<-c('a','b')
z1
class(z1)

w<-as.character(z)
w
class(w)

as.integer(w)

as.numeric(as.logical(c(5,0)))

z>1
as.numeric(z>1)

w<-c(1,"a")
w
class(w)


```


## Factors 
```{r}
ct<-c("jap","kor","sin","kor","jap","sin","sin")
class(ct)

fct<-factor(ct)
fct

levels(fct)

summary(fct)

table(fct)
```


## The function apply
  The functions apply, sapply and tapply are very useful
```{r}
income<-c(500,1200,4000,2300,2300,1234,1345)
income

tapply(income,fct,mean)
?tapply
?apply

med<-data.frame(patient=1:100,age=rnorm(100,mean=60,sd=12), treatment=gl(2,50,labels=c("Treatment","Control")))
med

tapply(med$age,med$treatment,mean)
```

END OF LECTURE 1: R BASICS

## Matrix and arrays

### Matrix operations

### Array operations


### Other operations with matrices

### Matrix multiplication

### Solutions of linear equations

### Eigen decomposition of a matrix
Find $\lambda$ (eigenvalues) such that $|A_\lambda I|=0$ and corresponding $\boldsymbol{x}$ (eigenvectors) such that $A\boldsymbol{x}=\lambda \boldsymbol{x}$.


## Lists 
List consists of an ordered collection of objects that can be of different or the same type.



## Dataframes

Dataframes are a tightly coupled collection of variables that share many of the properties of matrices and lists and is the fundamental data structure that will be used in most of this course.


## Tibbles and dyplr
.... not complete


END OF LECTURE ON R BASICS.